From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2017-11-01 paper.Rmd
WIP

The importance of $\tau_c$, or the degree of auto-correlation, which
emerges from our derivation above, has also been a subject of
significant interest in the study of stochastic population dynamics
[e.g.@Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015].
Early work rarely often assumed environmental noise would be
uncorrelated as a mathematical convenience.  As the derivation of
Eq \eqref{env_noise} makes clear, for a continuous-time process,
the relevant metric is the timescale of auto-correlation in the
environmental noise process relative to the timescale (auto-correlation)
of the population dynamics.  The Gaussian noise that emerges out
of the system-size approximation is always auto-correlated to some
degree -- if we measure our environment at consecutive times
infinitesimally far apart, we expect nearly identical values.  At
intervals much longer than the correlation time, consecutive
measurements are effectively independent: auto-correlated noise is
all about _relative_ timescales.

@@

A natural focus of much research into stochastic ecological systems
has been on understanding the implications for persistence of
individual and coexistence of interacting species. Demographic noise
in particular has fundamental implications for the long-term
equilibrium: any finite population, however big, is guaranteed to
eventually go extinct under demographic noise.  While extinction
is certainly a phenomenon that stands in stark contrast to the
long-term behavior of the corresponding deterministic equation,
this result is not particularly useful: "eventually" could be longer
than the existence of the planet. Rather, it is a reminder that we
must be careful about how we describe the 'long-term' behavior of
a model.  In place of the mathematical notion of a stationary
distribution to describe the long-term behavior of a stochastic
model, analysis of demographic stochasticity focuses on the concept
of quasi-stationary distributions, which in many cases correspond
more closely to the equilibrium of their deterministic counterparts
[@Ovaskainen2010; @Schreiber2017].

<!-- FIXME this section is still a fragment

[@Melbourne2008] Importance of demographic stochasticity and its interaction with environmental stochasticity in driving extinction risk. 
[@Bonsall2004] Markov chain environmental and demographic noise predator-prey metapopulation model.

@@

Insufficient space left for any treatment of these...

**Bet hedging**

@@

Noise can provide novel information in other ways as well.  Stochastic
fluctuations act like a continual miniature perturbation experiment,
constantly pushing a system away from an equilibrium to reveal its
behavior elsewhere in state space. Observing the size and
auto-correlation of these fluctuations thus provides insight on the
stability of an equilibrium point, (as can be seen, for instance
in the fluctuation-dissipation result we found for generic birth
death models).  This observation has been exploited with particular
success as a potential *early warning sign* that a system is loosing
stability; possibly due to an approaching catastrophic transition
or saddle-node bifurcation.  This pattern, known as critical slowing
down (in reference to the auto-correlation time in the physics
literature) was first proposed an an early warning sign for ecological
systems by @Wissel1984, but has received considerable empirical as
well as theoretical attention only more recently [@Scheffer2009;
@Scheffer2015].  Increases in two of these early warning indicators,
variance and auto-correlation, are illustrated along with the
timeseries of a simulation of the dynamics of yeast population
growth in Figure 4.  This simulation is based on a mechanistic model
developed by @Dai2012 that closely matches the population dynamics
they observed in serial dilutions of budding yeast _Saccharomyces
cervisiae_ on a sucrose medium. Because the relevant proteins are
located in the extra-cellular matrix rather than inside the cells,
budding yeast break down sucrose collaboratively, creating an Allee
effect below which the population is too small to sustain itself.
Increasing the serial dilution rate slowly over time drives the
stable population towards a fold bifurcation, which is accompanied
by the patterns of increasing variance and increasing auto-correlation
seen empirically in lab populations [@Dai2012] as well as the
numerical simulation shown here.

@@

<!-- @Holling1996 is responsible for the introduction of the unfortunate term "Engineering Resilience" to mean precisely the same thing as mathematicians and theoretical ecologists have refered to as 'stability' for decades.  Further confusion has been sown by the introduction of the energy-potential diagram into ecology, where the depth of the potential well is not a local property-->

@@

<!--
It is important to realize that such approaches can only ever measure local properties such as the return time in response to local perturbation (historically termed *stability*; which @Holling1996 landmark chapter unfortunately re-brands as "engineering resilience") and not global properties such as the size of the attractor basin / distance to the nearest tipping point (*ecological resilience*, or simply *resilience*).  This is illustrated rather vividly in @Dai2015, which illustrates how a system can lose resilience before losing stability, resulting in a scenario that provides less opportunity for early warning.

>>>>

The importance of $\tau_c$, or the degree of auto-correlation, which
emerges from our derivation above, has also been and continues to be a subject of
significant interest in the study of stochastic population dynamics
[e.g.@Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015].
Early work rarely often assumed environmental noise would be
uncorrelated as a mathematical convenience.  As the derivation of
Eq \eqref{env_noise} makes clear, for a continuous-time process,
the relevant metric is the timescale of auto-correlation in the
environmental noise process relative to the timescale (auto-correlation)
of the population dynamics.  The Gaussian noise that emerges out
of the system-size approximation is always auto-correlated to some
degree -- if we measure our environment at consecutive times
infinitesimally far apart, we expect nearly identical values.  At
intervals much longer than the correlation time, consecutive
measurements are effectively independent (white noise). 
The degree of auto-correlation in  noise is all about _relative_ timescales.

@@

A natural focus of much research into stochastic ecological systems
has been on understanding the implications for persistence of
individual and coexistence of interacting species. Demographic noise
in particular has fundamental implications for the long-term
equilibrium: any finite population, however big, is guaranteed to
eventually go extinct under demographic noise.  However, in most
models it is possible to define a notion of quasi-stationary 
distribution concentrated around the deterministic attractor,
and to prove the probability of extinction decreases exponentially with 
system size (e.g. see @Ovaskainen2010; @Schreiber2017 also reviews 
more recent literature which has established these results more generally). 
Generalizations about persistence and stable coexistence of interacting
species in stochastic models are difficult due to complex interactions
between demographic and environmental noise, population structure,
spatial heterogeneity, magnitude of noise, degree of autocorrelation and so forth
[e.g. @Higgins1997; @Coulson2001; @Melbourne2008; @Hart2016];
indeed care must be taken even in defining the approriate notion of 
persistence [e.g. @Schreiber2006].  Instead of these long-term limits,
here we will focus on a few examples of stochastic phenomena 
that can emerge on shorter time scales and provide some intuition for 
how the interaction between stochastic an nonlinear dynamics can 
create these surprising results.

@@

Insufficient space left for any treatment of these...
Permenance, robust permenance

**Bet hedging**, Lewontin & Cohen 1969; Levins 1969;

@@

Noise can provide novel information in other ways as well.  Stochastic
fluctuations act like a continual miniature perturbation experiment,
constantly pushing a system away from an equilibrium to reveal its
behavior elsewhere in state space. Observing the size and
auto-correlation of these fluctuations thus provides insight on the
stability of an equilibrium point, (as can be seen, for instance
in the fluctuation-dissipation result we found for generic birth
death models).  This observation has been exploited with particular
success as a potential *early warning sign* that a system is loosing
stability; possibly due to an approaching catastrophic transition
or saddle-node bifurcation.  This pattern, known as critical slowing
down (in reference to the auto-correlation time in the physics
literature) was first proposed an an early warning sign for ecological
systems by @Wissel1984, but has received considerable empirical as
well as theoretical attention only more recently [@Scheffer2009;
@Scheffer2015].  Increases in two of these early warning indicators,
variance and auto-correlation, are illustrated along with the
timeseries of a simulation of the dynamics of yeast population
growth in Figure 4.  This simulation is based on a mechanistic model
developed by @Dai2012 that closely matches the population dynamics
they observed in serial dilutions of budding yeast _Saccharomyces
cervisiae_ on a sucrose medium. Because the relevant proteins are
located in the extra-cellular matrix rather than inside the cells,
budding yeast break down sucrose collaboratively, creating an Allee
effect below which the population is too small to sustain itself.
Increasing the serial dilution rate slowly over time drives the
stable population towards a fold bifurcation, which is accompanied
by the patterns of increasing variance and increasing auto-correlation
seen empirically in lab populations [@Dai2012] as well as the
numerical simulation shown in Figure 4.

@@

<!-- FIXME Fragment, also tie back to flickering results, move some of that here-->

@@
